|

### 2021-03-10 05-Chp-CPRD-Analysis.Rmd
All working again after the update to Pandoc

```{block type='savequote', include=knitr::is_latex_output(), quote_author='(ref:cprd-quote)'}
When dealing with human beings controlled experiments frequently prove to be impracticable, so for a scientific basis for our assumptions we turn to past history to reconstruct the suspected causal chain of events - and then our statistical troubles may begin.
```
(ref:cprd-quote) --- Harold F. Dorn, 1953 [@dorn1953]

# Primary analysis of lipid regulating agents and dementia {#cprd-analysis-heading}
\minitoc <!-- this will include a mini table of contents-->

>>>>

```{block type='savequote', include=knitr::is_latex_output(), quote_author='(ref:cprdquote)', echo = TRUE}
When dealing with human beings controlled experiments frequently prove to be impracticable, so for a scientific basis for our assumptions we turn to past history to reconstruct the suspected causal chain of events - and then our statistical troubles may begin.
```
(ref:cprdquote) --- Harold F. Dorn, 1953 [@dorn1953]

# Primary analysis of lipid regulating agents and dementia {#cprd-analysis-heading}

\minitoc <!-- this will include a mini table of contents-->

### 2021-03-10 06-Chp-IPD-Analysis.Rmd
End of day: -3533
All working again after the update to Pandoc

* Description (including observation period, numbers, numbers with outcome, etc)
*
*

@@

>>>>

* Description (including observation period, numbers, numbers with outcome, etc)
* Whether they are a known genetically at-risk cohort
*

### Epic Norfolk

The European Prospective Investigation of Cancer - Norfolk is a  [@riboli1997; @riboli2002]

@@

Different approaches to combining subgroups. [@fisher2017]

## Discussion

Letters sent to

This is likely due to my junior position as a ECR

Range of reasons why

### 2021-03-10 front-and-back-matter/99-appendices.Rmd
End of day: -3533
All working again after the update to Pandoc

<!----------------------------------------------------------------------------->
# Other Appendix {#other-appendix-heading}

@@

This thesis was written in RMarkdown.
Several R packages were used as part of this project.`r gen_rmd_citation()`

## Producing risk-of-bias visualisation with `robvis` {#appendix-robvis}

@@

Risk of bias assessment - evaluation of the internal validity of studies included in a systematic review - often forms a key part of the evidence synthesis process, particularly in the health sciences.[@cochranechpt7] A well-developed family of tools is widely used, which have in common the characteristic that they evaluate specific domains of bias rather being constructed as a checklist or a quantitative score.[@cochranechpt7] These tools include the RoB 2 tool for randomized trials,[@sterne2019rob] the ROBINS-I tool for non-randomized studies of interventions,[@sterne2016robins] the QUADAS 2 tool for test accuracy and the ROBIS tool for systematic reviews.[@whiting2011quadas] Within each bias domains a judgement is reached about the strength of the study in that regard: for example, the first domain in the Cochrane RoB 2 tool deals with bias arising from the randomization process.[@sterne2019rob] Accessible graphics summarizing the results of these domain-based risk-of-bias assessments are included in reports of systematic reviews. A convenient plot in many reviews is a “traffic light” plot, which tabulates the judgement for each study in each domain. For larger numbers of studies, when such a table become unmanageable, a popular alternative is a weighted bar plot, which show the proportion of information with each judgement for each domain.[@higgins2008assessing]

Researchers can face a number of barriers in creating these plots. While some evidence synthesis platforms, such as Cochrane’s Review Manager,[@cochrane2014review] are able to produce these visualizations, not all researchers use these systems to conduct their systematic reviews, and copying the risk-of-bias data into these systems simply to produce the plots is inefficient and error prone. Likewise, creating the figures by hand, through software such as MS PowerPoint or Adobe Illustrator, may lead to unintentional errors and require the plots to be redrawn during an update to the review. Additionally, while the field of evidence synthesis software has grown rapidly in recent years,[@marshall2015systematic] this growth has not been equally distributed across the different aspects of the systematic review process. For example, a recent review found several software offerings aimed specifically at the abstract screening stage of the review process,[@harrison2020software] but no similar time- and error-reducing tool has been proposed for visualizing the results of risk-of-bias assessments.

@@

The tool is open-source and available to use free of charge. Users can download a stable version of the R package from CRAN ([https://cran.r-project.org/package=robvis](https://cran.r-project.org/package=robvis)); or access and contribute to the development version via GitHub ([https://github.com/mcguinlu/robvis](https://github.com/mcguinlu/robvis)).

@@

Development of `robvis` began in April 2019 at the Evidence Synthesis Hackathon (ESH), an event which brings together interested researchers, practitioners and coders to discuss and develop new open-source evidence synthesis technologies. Test versions of both the R package and the web app were made available in early June 2019, with attendees of the ESH and members of the Bristol Appraisal and Review of Research (BARR) group at the University of Bristol being invited to test the tool and provide feedback. This feedback, along with other feature suggestions from the wider evidence synthesis community captured via GitHub issues, was incorporated and the first release version of the package was uploaded to CRAN in November 2019. The tool has been well received and is beginning to be cited in the evidence synthesis literature.[@gibb2019consistent; @habadi2019prevalence; @veloso2020effectiveness;@simillis2020; @tanneru2020]

@@

-> NEED TO ADD DATA HERE!

&nbsp;

``` {r, eval = FALSE}

@@

``` {r trafficplot, echo = FALSE, eval = TRUE, fig.cap = "Example risk of bias traffic light plot created using `robvis`", out.width = "100%", fig.align = "centre"}

@@

``` {r summaryplot, echo = FALSE, eval = TRUE, fig.cap = "Example risk of bias summary plot created using `robvis` and the example ROB2 dataset", out.width = "100%"}

@@

``` {r robvisarguments, echo = FALSE, eval = TRUE}

@@

robvis_downloads <- plyr::round_any(sum(robvisstats$count),100, f = floor)

@@

As of `r date`, `robvis` has been downloaded more than `r robvis_downloads` times. It has been well recieved but the systematic review community, and has been cited frequently in the published literature. A paper describing the tool was published in a special issue of Research Synthesis Methods focusing on data visualisation methods. A chapter on the tool has been incorporated in to the “Doing Meta-Analysis in R” online textbook.[@mathias_harrer_2019_2551803]

>>>>

<!----------------------------------------------------------------------------->

# Other Appendix {#other-appendix-heading}

@@

This thesis was written in RMarkdown. Several R packages were used as part of this project.[@base; @cranlogs; @dplyr; @flextable; @ggplot2; @glue; @gt; @here; @kableExtra; @knitr; @medrxivr; @plyr; @robvis; @stringr; @tidyverse; @wordcountaddin; @xfun; @rbiorxiv]

## Producing risk-of-bias visualisations with `robvis` {#appendix-robvis}

@@

Risk of bias assessment - evaluation of the internal validity of studies included in a systematic review - often forms a key part of the evidence synthesis process, particularly in the health sciences.[@cochranechpt7] A well-developed family of tools is widely used, which have in common the characteristic that they evaluate specific domains of bias rather being constructed as a checklist or a quantitative score.[@cochranechpt7] These tools include the RoB 2 tool for randomized trials,[@sterne2019rob] the ROBINS-I tool for non-randomized studies of interventions,[@sterne2016robins] the QUADAS 2 tool for test accuracy and the ROBIS tool for systematic reviews.[@whiting2011quadas] Within each bias domains a judgement is reached about the strength of the study in that regard: for example, the first domain in the Cochrane RoB 2 tool deals with bias arising from the randomization process.[@sterne2019rob] Accessible graphics summarizing the results of these domain-based risk-of-bias assessments are included in reports of systematic reviews. A convenient plot in many reviews is a "traffic light" plot, which tabulates the judgement for each study in each domain. For larger numbers of studies, when such a table become unmanageable, a popular alternative is a weighted bar plot, which show the proportion of information with each judgement for each domain.[@higgins2008assessing]

Researchers can face a number of barriers in creating these plots. While some evidence synthesis platforms, such as Cochrane's Review Manager,[@cochrane2014review] are able to produce these visualizations, not all researchers use these systems to conduct their systematic reviews, and copying the risk-of-bias data into these systems simply to produce the plots is inefficient and error prone. Likewise, creating the figures by hand, through software such as MS PowerPoint or Adobe Illustrator, may lead to unintentional errors and require the plots to be redrawn during an update to the review. Additionally, while the field of evidence synthesis software has grown rapidly in recent years,[@marshall2015systematic] this growth has not been equally distributed across the different aspects of the systematic review process. For example, a recent review found several software offerings aimed specifically at the abstract screening stage of the review process,[@harrison2020software] but no similar time- and error-reducing tool has been proposed for visualizing the results of risk-of-bias assessments.

@@

The tool is open-source and available to use free of charge. Users can download a stable version of the R package from CRAN (<https://cran.r-project.org/package=robvis>); or access and contribute to the development version via GitHub (<https://github.com/mcguinlu/robvis>).

@@

Development of `robvis` began in April 2019 at the Evidence Synthesis Hackathon (ESH), an event which brings together interested researchers, practitioners and coders to discuss and develop new open-source evidence synthesis technologies. Test versions of both the R package and the web app were made available in early June 2019, with attendees of the ESH and members of the Bristol Appraisal and Review of Research (BARR) group at the University of Bristol being invited to test the tool and provide feedback. This feedback, along with other feature suggestions from the wider evidence synthesis community captured via GitHub issues, was incorporated and the first release version of the package was uploaded to CRAN in November 2019. The tool has been well received and is beginning to be cited in the evidence synthesis literature.[@gibb2019consistent; @habadi2019prevalence; @veloso2020effectiveness; @simillis2020; @tanneru2020]

@@

-\> NEED TO ADD DATA HERE!

@@

```{r trafficplot, echo = FALSE, eval = TRUE, fig.cap = "Example risk of bias traffic light plot created using `robvis`", out.width = "100%", fig.align = "centre"}

@@

```{r summaryplot, echo = FALSE, eval = TRUE, fig.cap = "Example risk of bias summary plot created using `robvis` and the example ROB2 dataset", out.width = "100%"}

@@

```{r robvisarguments, echo = FALSE, eval = TRUE}

@@

robvis_downloads <- format(plyr::round_any(sum(robvisstats$count),100, f = floor), scientific = FALSE)

@@

As of `r date`, `robvis` has been downloaded more than `r robvis_downloads` times. It has been well received but the systematic review community, and has been cited frequently in the published literature. A paper describing the tool was published in a special issue of Research Synthesis Methods focusing on data visualisation methods. A chapter on the tool has been incorporated in to the "Doing Meta-Analysis in R" online textbook.[@mathias_harrer_2019_2551803]

### 2021-03-10 index.Rmd
All working again after the update to Pandoc

#######################
## bibliography path ##
#######################
bibliography: 
  - bibliography/references.bib
  - bibliography/pkg-refs.bib
csl: bibliography/nature.csl
bibliography-heading-in-pdf: Bibliography

>>>>

#######################
## bibliography path ##
#######################
bibliography: 
    - "bibliography/references.bib"
csl: bibliography/nature.csl
bibliography-heading-in-pdf: Bibliography

### 2021-03-10 reviews/year_2/report.Rmd
All working again after the update to Pandoc

x = 0
```

@@

**Title**:

@@

**Current date**: 19/01/2021

**PhD Start date**: 01/01/2019

@@

# Overview of PhD
<!-- ## Rationale -->

<!-- Several primary studies and systematic reviews exist that seek to examine the relationship between blood lipid levels, or interventions which modify blood lipid levels such as statins, and the subsequent risk of dementia. However, as yet, there has been no systematic attempt to find and integrate all available sources of evidence from distinct study designs to address this research question using a triangulation framework.[@lawlor2016a] -->

<!-- This thesis aims to find, critically appraise and synthesize the available evidence related to this research question, and then integrate this with the primary evidence produced as part of this thesis. -->

<!-- &nbsp; -->

## Aim {#aim}
The central aim of this thesis is to: <br>

* **To investigate the relationship between blood lipid levels, and interventions that modify blood lipid levels, and dementia outcomes**

@@

## Objectives {#objectives}
To achieve this aim, there are several smaller objectives, which map to the proposed chapters of my PhD outline in Section \@ref(chapter-progress):

1. Identify all existing on the relationship between blood lipid levels (and interventions that affect blood lipid levels) and dementia, regardless of study design, and including studies available solely as a preprint.

@@

4. Integrate the existing evidence from the review with the primary evidence produced as part of this thesis as part of a qualitative triangulation exercise.

## Theoretical framework
The main theoretical framework used in this thesis is evidence synthesis - the discovery and critical integration of all available evidence on a research question in order to either: a) provide a more definitive answer to that question or; b) highlight gaps in the existing evidence base, so that future research can address questions that have yet to be answered, or can re-examine existing questions in a way that increases our confidence in the result.

The primary evidence generation aspects of this thesis are performed with the primary intention of providing a further source of evidence for the evidence synthesis/triangulation aspect.

@@

The thesis is laid out in a similar manner to that successfully defended by [Sam Abbott of the MRC IEU](https://www.samabbott.co.uk/thesis). Each chapter is self-contained in that it presents the relevant methods and results for that particular analysis, and so there are no dedicated methods chapters.

@@

**Include Gantt chart here**

@@

This chapter introduces a tool built in R to allow for systematic searching of the [medRxiv](https://www.medrxiv.org/) and [bioRxiv](https://www.biorxiv.org/) preprint repositories.

Preprints represent an important source of grey literature and so should be included as part of a systematic review, and searching these repositories. However, the native search interface on the preprint websites do not allow for complex search strategies (e.g. using Boolean operators or wildcards) and are not transparent or repeatable, with the same search producing wildly different numbers of records over time.

To allow for easy systematic searching of this literature souce as part of the systematic review described in Chapter 4, the `medrxivr` tool was developed.

@@

The code underlying the tool underwent stringent peer review via the rOpenSci review process (you can see the discussion between myself and the reviewers [here](https://github.com/ropensci/software-review/issues/380)), following which a short paper describing the tool was [published in the Journal of Open Source Software](https://joss.theoj.org/papers/10.21105/joss.02651).

This chapter represents an expanded version of the published paper, expanding on the motivation, key functionality, reception and planned future development of the tool. A  draft of this chapter is complete.

@@

__Note:__ evidence synthesis software development has been a key component of my work during my Fellowship. At my last APR, Tom Gaunt flagged in his comments that I should consider including some of this work as a chapter in my thesis, if I could tie it in with the thesis as whole. Based on this, I am keen to include the medrxivr tool, as it was designed and developed specifically to help with the review described in Chapter 4. However, I would be very glad to hear your opinion on whether the inclusion of this work as a chapter makes sense in terms of the thesis as a whole.

@@

![](../../figures/sys-rev-tools/medrxiv-role.png)

@@

This chapter presents the methods and findings of a comprehensive systematic review into the relationship between blood lipids levels/statins and dementia subtypes. The review included all types of study design including randomised controlled trials, Mendelian randomisation analyses, and non-randomised studies of exposures and interventions.

@@

* Literature search (100%)
* Title and abstract/Full-text screening (100%)
* Data extraction (100%)
* Risk of bias assessment (75%)
* Writing (50%)

@@

![](../../figures/sys-rev/prisma_flow.png)

@@

The review . Using other published reviews to verify the results demonstrated that the search was sufficiently comprehensive to capture all previously identified studies, plus some studies that had not been included in any previous review.

Data extraction has been completed for the 1-5 studies included in the review and

As dual screening and extraction is recommended for systematic reviews, this element of my thesis has involved a number of external researchers. I have found the people-management aspect of the review challenging, but it has provided good experience in terms of leading a review team.

For space, and to illustrate the proposed triangulation below

@@

One of the take-aways from my last APR was to make the "added value" of this work beyond existing review clear. I believe there are two aspects where this review adds value above what is already available in the published literature:

* _Comprehensiveness:_ While several reviews of this research topic exist,[@chu2018b; @yang2020; @muangpaisan2010; @poly2020b] the overlap between the list of studies included in each is not 100%. As part of this review, I have not only performed a original search of primary literature databases, but have also searched the reference lists of comparable reviews to ensure no study has been omitted.
* _Structured risk of bias assessment:_ The majority of the highly cited reviews on this topic either do not formally consider the risk of bias in the observational studies they include or do not use an appropriate domain-based assessment tool (e.g.ROBINS-I/E). This is important area in which this thesis can add value, as based on the risk of bias assessments I have performed to date, several primary studies are at high risk of bias and this should be reflected in the findings of any  review on this topic.

@@

<!----------------------------------------------------------------->
## Chapter 5: Observational analyses

@@

Rather than reproduce the methods/findings here, I have attached the initial draft of the paper this chapter is based on at the end of this report.

@@

In relation to the results presented in the attached paper, there are two key discussion points. The first is that while statins are associated with a slight reduction in the risk of probable AD (HR: 0.91, 95%CI: 0.88-0.95) and possible AD (HR: 0.95, 95%CI: 0.92-0.99), we still see an increased risk of vascular dementia (HR: 1.7, 95%CI: 1.62-1.77) and other dementia (HR: 1.7, 95%CI: 1.62-1.77) in those using statins. This reflects the initial findings of this analysis presented at the last APR, and was not attenuated even after using multiple imputation to account for potential selection bias caused by the original complete case analysis and accounting for several potentially important confounders.

Secondly, following a recommendation by a co-author to use the same analysis strategy with a known control (coronary heart disease; CHD), we found that statins appeared to increase the risk of CHD, in contrast to the the results of large RCTs. This finding suggests the presence of substantial confounding by indication, in that those who go on to take a statin are fundamentally different from those who do not, and has been replicated in other standard mulitvariable HR analyses.[@danaei2013a] A suggested approach to frame this (possible insurmountable) issue has been to rewrite the paper as an example of the dangers of using EHR in cases where strong confounding by indication is likely. However, it would be particularly useful to get reviewers feedback on any further analytical strategies we could attempt to address this issue.

In terms of the broader thesis, while this limitation is important, it presents a good opportunity to compare the effect of different analytical strategies with competing directions of bias (immortal time bias makes statins look better, confounding by indication makes statins look worse) of bias on the same underlying question.

Finally, in an attempt to assess whether these (likely spurious) findings for vascular dementia are a product of the CPRD data or our analytical strategy, we recently attempted to replicate an analysis performed by Smeeth et al. in the THIN EHR database which found a protective effect of statins on dementia risk (HR 0.81, 99% CI 0.69-0.96).[@smeeth2009a] However, following correspondence with the author team to acquire the code-list they used to define dementia cases, we confirmed that it was substantially different from that used in our study. This fact, coupled with concerns over their analytical strategy (which seemed to adjust for covariates on the causal pathway), limited the value of this attempted replication.

@@

Similar to the systematic review, feedback from my last APR suggested that the "added value" of this study above those that already exist should be made clearer.

There are two particular:
* _Size of the CPRD and length of follow-up:_ Having reviewed the other studies included in the review, this CPRD analysis would be one of the largest studies of this research question.
* _Addressing the limitations of other observational analyses:_ Analysing this data has provided the opportunity to use a separate analytical technique to many of the studies identified in the systematic review, As an example, the Hippsley-Cox BMJ paper examining the effect of statins, which makes use of the THIN EHR database, likely suffers from immortal time bias as exposed and unexposed participants are not followed up from a common time point.[@hippisley-cox2010] As touched on in the section above, this provides an additional evidence point with a different source and direction of bias, which is useful for the triangulation aspect of the thesis.

@@

This chapter will draw together different sources of information (including the primary analysis performed as part of this thesis) in a triangulation framework, along with a consideration of the key sources/direction of bias in each, to inform.

As flagged in my last APR, this chapter will make clear that "triangulation" in the context of my thesis is a qualitative integration of the distinct evidence sources available (systematic review, CPRD study, IPD meta-analysis).

@@

However, in the coming year, I plan to attend the "Mendelian randomisation" and "Advanced methods for multiple imputation" short courses at the University of Bristol, in addition to the "Causal Inference in Epidemiology: Recent Methodological Developments" short course at LSHTM and the "Writing about your research" course at the Royal Society.

@@

In the review period, I presented on the preprint search tool described in Chapter 3 at the Bristol  "Autosynthesis Club" and "Methods in Evidence Synthesis" seminars.

It's only in writing this report that I realised that I am yet to present on the other parts of my thesis, and so this is a particular aspect of my PhD that I would like to develop further in the coming year. As such, I aim to present at more diverse events in the coming year, including at an IEU monthly meeting and the Faculty of Health Sciences research showcase.

@@

* MSc Epidemiology/Public Health "Clinical Epidemiology" module
    * Tutored a small group of several weeks
    * Marked the end-of-module assessments
* Introduction to R short course for new PhD students, Bristol Medical School short course
    * Presented a lecture on advanced R topics (data visualisation, literate programming and web applications)
* Introduction to data visualisation and web applications using R, Bristol Medical School short course
    * Designed and delivered the web applications aspects of the course
    * Tutor helping on practicals on data visualisations and literate programming with Rmarkdown
* Drop-in "Intro to R" sessions
    * Organised and ran an hour long session each Friday in the autumn term to help answer new PGR students' questions about R.

I have also submitted a proposal to supervise an MSc student for the summer term. The project aims to examine the overlap between the set of primary studies included in the several existing systematic reviews of blood lipid levels and dementia (as mentioned in the description of Chapter 4, above).

@@

* *Comparison of data availability statements:*
Using medrxivr (the preprint search tool described in Chapter 2), myself and a fellow PhD student compared the availability of data for the same manuscript when published under two different data-sharing policies (i.e. that of the preprint repository vs. the peer-reviewed journal). The preprint of the study is available [here](https://doi.org/10.31222/osf.io/p75xe).

* *Making primary research synthesis-ready:* I acted as last author on a commentary describing the best practice and advatnages for primary researchers in the field of prevention science to ensure that their research is "synthesis-ready" (i.e. can be easily found and integrated into an evidence synthesis project). This commentary is under consideration at "Prevention Science" and a preprint is available [here](https://doi.org/10.31222/osf.io/ptg9j).

>>>>

x = 0
y = 0
```

@@

**Title**: Investigating the relationship between blood lipids levels, and interventions that modify blood lipid levels, and dementia outcomes.

@@

**PhD start date**: 01/01/2019

@@

# Overview of PhD

## Aim {#aim}
The central aim of this thesis is: <br>

* **To investigate the relationship between blood lipid levels, and interventions that modify blood lipid levels, and dementia outcomes.**

@@

## Objectives {#objectives}
To achieve this aim, there are several smaller objectives, which map to the proposed chapters of my thesis outlined in Section \@ref(chapter-progress):

1. Identify all existing evidence on the effect of blood lipid levels (and interventions that affect blood lipid levels) on dementia risk, regardless of study design and including studies available solely as a preprint.

4. Integrate the existing evidence identified by the review with the primary evidence produced as part of this thesis in a qualitative triangulation exercise.

@@

## Theoretical framework
The main theoretical framework used in this thesis is evidence synthesis - the identification, critical assessment, and integration of all available evidence on a research question in order to either: a) provide a more definitive answer to that question; or b) highlight gaps in the existing evidence base.

While this thesis does include a primary evidence generation element (Chapter 5 - observational analysis of CPRD data), this was performed with the intention of providing a further source of evidence for the evidence synthesis/triangulation aspects.

@@

The thesis is laid out in a similar manner to that successfully defended by [Sam Abbott of the MRC IEU](https://www.samabbott.co.uk/thesis). Each chapter is self-contained, presenting the relevant methods and results for that particular analysis, and so there are no dedicated methods chapters.

@@

![](../../figures/gantt.png){width=450}
&nbsp;

__Figure `r (x <- x + 1)`: Gantt chart showing research milestones, training courses and conferences for the last 12 months of my PhD thesis.__ 
&nbsp;
&nbsp;

@@

This chapter introduces a tool built in R to allow for systematic searching of the [medRxiv](https://www.medrxiv.org/) and [bioRxiv](https://www.biorxiv.org/) preprint repositories. Preprints represent an important source of grey literature and so should be included as part of a systematic review. However, the native search interface on preprint websites do not allow for complex search strategies (e.g. those using Boolean operators or wildcards) and are not transparent or reproducible, with the same search producing wildly different numbers of records over time.

To allow for easy systematic searching of this literature source as part of the systematic review described in Chapter 4, the `medrxivr` tool was developed.

__Note:__ Evidence synthesis software development has been a key component of my work during my Fellowship. At my last progress review, Tom Gaunt (PGR Director) flagged in his comments that I should consider including some of this work as a chapter in my thesis, if I could tie it in with the thesis as whole. Based on this, I am keen to include the medrxivr tool, as it was designed and developed specifically to help with the review described in Chapter 4. However, I would be very glad to hear your opinion on whether the inclusion of this work as a chapter makes sense in terms of the thesis as a whole.

@@

The code underlying the tool underwent peer review via the rOpenSci review process (you can see the discussion between myself and the reviewers [here](https://github.com/ropensci/software-review/issues/380)), following which a short paper describing the tool was [published in the Journal of Open Source Software](https://joss.theoj.org/papers/10.21105/joss.02651).

This chapter represents an expanded version of the published paper, discussing the motivation, key functionality, reception and planned future development of the tool. A draft of this chapter is complete.

@@

@@

![](../../figures/sys-rev-tools/medrxiv-role.png){width=350}

@@

This chapter will present the methods and findings of a comprehensive systematic review into the relationship between blood lipids levels/blood lipids interventions (e.g. statins) and dementia. The review included all types of study design including randomised controlled trials, Mendelian randomization analyses, and non-randomised studies of exposures and interventions (NRSI/NRSE).

@@

In terms of the different stages of the review, progress currently stands as:

* Literature search: 100%
* Title and abstract/Full-text screening: 100%
* Data extraction: 100%
* Risk of bias assessment: 75%
* Analysis: 60%
* Writing: 50%

@@

![](../../figures/sys-rev/prisma_flow.png){width=350}

@@

First, in response to issues flagged at last year's progress review (comprehensiveness of search/is screening a 10% sample in duplicate sufficient if inter-rater reliability is low), searching the bibliographies of published reviews on this topic demonstrated that the search was sufficiently comprehensive to capture all previously identified studies in addition some studies that had not been included in any previous review. Additionally, inter-rater agreement on a random 10% subsample of records was high for both the title/abstract and full-text screening stages of the review.

For the sake of space, the full methods and results of the review are not presented here, but the protocol is [available online](https://doi.org/10.17605/OSF.IO/VTW5Y). As a short summary, 127 reports describing 105 primary studies were included in the review (see the PRISMA flow diagram above for an overview of the movement of records through the review). The majority of studies identified by the review are non-randomised studies, with only two randomised controlled trails and eight Mendelian randomization studies identified. Data extraction has been completed for all studies, and risk-of-bias assessments for each are almost complete.

To enable comparison with the results of the observational analysis presented as part of Chapter 5, the results of a meta-analysis examining the effect of statins on dementia outcomes in included NRSI is presented. Statin use was associated with a reduced risk of all-cause dementia (HR: 0.86, 95%CI 0.83-0.88) and Alzheimer's disease (HR: 0.78, 95%CI 0.69-0.89), but were not found to have an effect on the risk of vascular dementia (HR: 0.93, 95%CI 0.76-1.14).

![](../../figures/forestplot_acd.png){width=350}

__Figure `r (x <- x + 1)` - Results of a random-effects meta-analysis of non-randomised studies examining the effect of statin use on the risk of all-cause dementia.__

![](../../figures/forestplot_ad.png){width=350}

__Figure `r (x <- x + 1)` - Results of a random-effects meta-analysis of non-randomised studies examining the effect of statin use on the risk of Alzheimer's disease.__

![](../../figures/forestplot_vad.png){width=350}

__Figure `r (x <- x + 1)` - Results of a random-effects meta-analysis of non-randomised studies examining the effect of statin use on the risk of vascular dementia.__

In terms of the meta-analysis of the non-intervention studies, this was complicated by the different ways in which lipid fractions were quantified in different studies (e.g. grouped based on clinical guidelines or quantiles, or dichotomized into a hypercholesterolemia variable). However, following discussion with my supervisors who pointed me towards useful material on this subject, this element of the analysis is almost complete.

One interesting discussion point highlighted by the review is the absence of data on vascular dementia as an outcome. As illustrated in the meta-analysis above, there were substantially fewer non-randomised studies reporting vascular dementia as an outcome compared with those reporting all-cause dementia or Alzheimer's disease outcomes. This is reflected in the available randomised controlled trials and Mendelian randomization studies (as no genetic consortia data are available for the vascular dementia outcome, MR studies of this condition are precluded). This will restrict the triangulation exercise for this outcome, as it will have to rely on different analytical strategies within a single study design (non-randomised studies) rather than incorporating fundamentally different study designs.

A final point on the progress of the review is that, due to the need for dual screening and data extraction, a number of external researchers became involved in this review. I found the people-management aspect particularly challenging and could definitely have improved the process through better communication of deadlines, but it has provided good experience of leading a review team.

@@

One of the take-aways from my last progress review was to make the "added value" of this work beyond existing reviews clear. I believe there are three aspects where this review is distinct from those reviews already available in the published literature:

* _Comprehensiveness:_ While several reviews of this research topic exist,[@chu2018b; @yang2020; @muangpaisan2010; @poly2020b] the overlap between the list of studies included in each is not 100%. As part of this review, I have not only performed a original search of primary literature databases, but have also screened the reference lists of comparable reviews to ensure no study has been omitted.
* _Structured risk of bias assessment:_ The majority of the highly cited reviews on this topic either do not formally consider the risk of bias in the observational studies they include or do not use an appropriate domain-based assessment tool (e.g. ROBINS-I/E). This is important area in which this thesis can add value, as based on the risk-of-bias assessments I have performed to date, several primary studies are at high risk of bias and this should be reflected in the findings of any  review on this topic.
* _Inclusion of preprints:_ Unlike other available reviews and enabled by the tool described in Chapter 3, this review systematically searched  preprinted health-related manuscripts as a source of grey literature. As part of this chapter, I plan to examine the extent of the additional information provided to the review by the inclusion of preprints.

@@

<!----------------------------------------------------------------->
## Chapter 5: Observational analysis

@@

Rather than reproduce the methods/findings of this chapter here, I have attached the initial draft of the paper this chapter is based on at the end of this report.

@@

In relation to the results presented in the attached paper, there are two key discussion points. The first is that while statins are associated with a slight reduction in the risk of probable (HR: 0.91, 95%CI: 0.88-0.95) and possible Alzheimer's Disease (HR: 0.95, 95%CI: 0.92-0.99), there is an increased risk of vascular dementia (HR: 1.7, 95%CI: 1.62-1.77) and other dementia (HR: 1.13, 95%CI: 1.09-1.18) in those using statins. This reflects the initial findings of this analysis presented at the last progress review, and was not attenuated even after using multiple imputation to account for potential selection bias caused by the original complete case analysis and after accounting for several potentially important confounders.

Secondly, following circulation of the draft, a co-author recommended that I use the same analysis strategy with a known control (coronary heart disease; CHD). However, when running this analysis, I found that statins appeared to increase the risk of CHD, in contrast to the results of published large RCTs. This finding suggests the presence of substantial confounding by indication in this analysis, in that those who go on to take a statin are fundamentally different from those who do not, and this has been seen in other standard multivariable HR analyses of the effect of statins on CHD.[@danaei2013a] A suggested approach to frame this (possible insurmountable) issue has been to rewrite the paper as an example of the dangers of using EHR in cases where strong confounding by indication is likely. However, it would be particularly useful to get reviewers feedback on any further analytical strategies I could attempt to address this issue.

In terms of the broader thesis, while these concerns mean I have low confidence in the accuracy of the result, it presents a good opportunity for comparison with other published papers that used different analytical strategies where the source and direction of bias is different (e.g. confounding by indication in our analysis likely makes statin use look worse but immortal time bias in other papers likely makes statin use look better).

Finally, in an attempt to assess whether these unexpected findings for vascular dementia are a product of the CPRD data or our analytical strategy, I recently attempted to replicate an analysis performed by Smeeth et al. which used the THIN EHR database and found a protective effect of statins on dementia risk (HR 0.81, 99% CI 0.69-0.96).[@smeeth2009a] However, following correspondence with the author team to acquire the code-list they used to define dementia cases, I confirmed that it was substantially different from that used in our study. This fact, coupled with concerns over their analytical strategy (which seemed to adjust for covariates on the causal pathway), limited the value of this attempted replication.

@@

Similar to the systematic review presented in Chapter 4, feedback from my last progress review suggested that the "added value" of this study above those that already exist should be made clearer. There are two particular motivations for this analysis above those already available in the published literature:

* _Size of the CPRD and length of follow-up:_ Having reviewed the other studies identified by the systematic review in Chapter 4, this analysis of 1.7 million participants is one of the largest studies of this research question.
* _Addressing the limitations of other observational analyses:_ Analyzing this data has provided the opportunity to use a separate analytical technique to many of the studies identified in the systematic review, As an example, the Hippsley-Cox BMJ paper examining the effect of statins, which makes use of the THIN EHR database, likely suffers from immortal time bias as exposed and unexposed participants are not followed up from a common time point.[@hippisley-cox2010] As touched on in the section above, this provides an additional evidence point with a different source and direction of bias, which is useful for the triangulation aspect of the thesis.

@@

This chapter will draw together different sources of information (including the primary analysis performed as part of this thesis), along with a consideration of the key sources/direction of bias in each, in a qualitative triangulation framework.[@lawlor2016a]

As flagged in my last progress review, this chapter will make very clear that "triangulation" in the context of my thesis is a qualitative integration of the distinct evidence sources available (systematic review, CPRD study, IPD meta-analysis).

@@

However, in the coming year, I have secured places on the  "Mendelian randomization" and "Advanced methods for multiple imputation" short courses at the University of Bristol, in addition to the "Causal Inference in Epidemiology: Recent Methodological Developments" short course at LSHTM and the "Writing about your research" course at the Royal Society.

@@

In the review period, I presented on the preprint search tool described in Chapter 3 at the Bristol  "Auto-synthesis Club" and "Methods in Evidence Synthesis" seminars.

It's only in writing this report that I realized that I am yet to present on the other parts of my thesis, and so this is a particular aspect of my PhD that I would like to develop further in the coming year. As such, I aim to present at more diverse events in the coming year, including at an IEU monthly meeting and the Faculty of Health Sciences research showcase.

@@

* MSc Epidemiology/Public Health "Clinical Epidemiology" module
    * Tutored a small group for several weeks
    * Marked the end-of-module assessments
* "Introduction to R" short course for new PhD students, Bristol Medical School short course
    * Presented a lecture on advanced R topics (data visualisation, literate programming and web applications)
* "Introduction to Data Visualisation and Web Applications using R", Bristol Medical School short course
    * Designed and delivered the web applications aspects of the course
    * Tutor helping on practicals on data visualisations and literate programming with Rmarkdown
* Drop-in "Intro to R" sessions
    * Organised and ran an hour long session each Friday in the autumn term to help answer new PGR students' questions about R.

I have also had a proposal accepted to supervise an MSc student for the summer term. The project aims to examine the overlap between the set of primary studies included in the several existing systematic reviews of blood lipid levels and dementia (as mentioned in the description of Chapter 4, above).

@@

* *Comparison of data availability statements:*
Using medrxivr (the preprint search tool described in Chapter 2), fellow PhD student and I compared the availability of data for the same manuscript when published under two different data-sharing policies (i.e. that of the preprint repository vs. the peer-reviewed journal). The preprint of the study is available [here](https://doi.org/10.31222/osf.io/p75xe).

* *Making primary research synthesis-ready:* I am last author on a commentary describing the best practice and advantages for primary researchers in the field of prevention science to ensure that their research is "synthesis-ready" (i.e. can be easily found and integrated into an evidence synthesis project). This commentary is under consideration at "Prevention Science" and a preprint is available [here](https://doi.org/10.31222/osf.io/ptg9j).
